---
title: 'NLP analysis: detecting frequency jumps'
author: "Andrey Anikin"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 8
    fig_height: 6
    code_folding: show
---

```{r echo=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r echo=FALSE}
library(soundgen)
library(ggplot2)
```


# A single synthetic sound

Here is what happens when pitch tracking is imperfect - we think there are two extra frequency jumps:

```{r}
# png('../pix/pitchJump_validate_ex_spec.png', width = 15, height = 12, units = 'cm', res = 600)
detectNLP('../sounds/beyond_speech/nonverbal_cry_F_482_ut_sad-cry_34-f.wav', 
          plot = TRUE, ylim = c(0, 3))
# dev.off()
```

 Or, to make everything look neat & nice with correct pitch tracking:
 
```{r}
detectNLP('../sounds/beyond_speech/nonverbal_cry_F_482_ut_sad-cry_34-f.wav', 
          pitchManual = '../data/pitch_manual_1745.csv',
          plot = TRUE, ylim = c(0, 3))
```
 


# Optimization

Optimize based on manually annotated pitch jumps in 1518 human vocalizations

## Preparation

First we load the results of acoustic analysis and annotated pitch jumps in these 1518 vocalizations.

```{r class.source = 'fold-hide'}
df = readRDS('../data/bs_per_frame.RDS')
# or w/o manual pitch tracking:
# df = readRDS('../data/bs_per_frame_autom.RDS')

## optimize the detection of pitch jumps
table(df$fj)
step = df$time[2] - df$time[1]
# find the time of pitch jumps (upsampling put multiple frames together as pj)
all_ann = read.csv('../data/nlp_manual.csv')
all_ann = all_ann[all_ann$label == 'fj', ]
nrow(all_ann)  # 130 freq jumps
length(unique(all_ann$file))  # 65 recordings contain at least 1 freq jump
```

So we have 65 recordings with at least annotated one frequency jump, for a total of 130 jumps. Since these jumps are supposed to be instantaneous, let's recalculate their time as the mean of the annotated from-to times (duration in meaningless here - it's just an artifact of annotation).

```{r}
pjs = data.frame(file = all_ann$file, time = rowMeans(all_ann[, c('from', 'to')]))
summary(pjs)
```

Now we need a function that detects frequency jumps automatically from f0 contours - that's soundgen:::findJumps - and compares the results with manual annotations. To make sure we strike a reasonable balance between seeing frequency jumps everywhere (a lot of false alarms) and being too conservative and missing real jumps (a lot of false negatives), F1-score is used as the accuracy metric.

```{r class.source = 'fold-hide'}
check_pj = function(jumpThres = 14, jumpWindow = 100, tol_ms = 50, plot = FALSE) {
  files = unique(df$file)
  out = data.frame(file = files, n_manual = 0, n_found = 0, TP = 0, FP = 0)
  for (i in 1:length(files)) {
    f = files[i]
    idx_i = which(df$file == f)
    pj_i = soundgen:::findJumps(pitch = df$pitch[idx_i], step = step, 
                                jumpThres = jumpThres, jumpWindow = jumpWindow)
    time_pj_i = which(pj_i) * step - step/2
    n_pj_i = length(time_pj_i)
    out$n_found[i] = n_pj_i
    if (f %in% pjs$file) {
      # there are some manually annotated pitch jumps in this file
      times_true = pjs$time[which(pjs$file == f)]
      out$n_manual[i] = length(times_true)
      if (n_pj_i > 0) {
        # we found some pitch jumps
        # true positives: correctly detected pj's within tol_ms
        out$TP[i] = sum(apply(as.matrix(times_true), 1, function(x) any(abs(x - time_pj_i) <= tol_ms)))
        # false positives
        out$FP[i] = sum(apply(as.matrix(time_pj_i), 1, function(x) !any(abs(x - times_true) <= tol_ms)))
      } else {
        # we didn't find any pitch jumps, but there are in fact some
        out$TP[i] = 0
        out$FP[i] = 0
      }
    } else {
      # there are no manually annotated pitch jumps in this file
      times_true = numeric(0)
      if (n_pj_i > 0) {
        # we have hallucinated some pitch jumps where there are none
        out$TP[i] = 0
        out$FP[i] = n_pj_i
      } else {
        # no jumps present, none found
        out$TP[i] = 0
        out$FP[i] = 0
      }
    }
    
    if (plot & ((n_pj_i > 0) | (length(times_true) > 0))) {
      plot(step * (1:length(df$pitch[idx_i])), df$pitch[idx_i], type = 'b', 
           pch = 16, xlab = 'Time, ms', ylab = 'f0, Hz', main = f)
      abline(v = times_true, lty = 2, col = 'blue')
      y1 = min(df$pitch[idx_i], na.rm = TRUE) + diff(range(df$pitch[idx_i], na.rm = TRUE)) / 5
      for (tj in time_pj_i) 
        arrows(x0 = tj, x1 = tj, y0 = 0, y1 = y1, lwd = 2, length = .05, col = 'purple')
      invisible(readline(prompt="Press [enter] to continue")) 
    }
  }
  
  # calculate F1-score
  sums = colSums(out[, 2:5])
  precision = as.numeric(sums['TP'] / sums['n_found'])
  recall = as.numeric(sums['TP'] / sums['n_manual'])
  F1 = 2 / (1/precision + 1/recall)
  return(list(out = out, F1 = F1, precision = precision, recall = recall))
}
```

Sanity checks - let's look at the distribution of TP/FP:

```{r}
temp = check_pj()
temp$F1
table(temp$out$TP, temp$out$FP)  # confusion matrix with default settings
```


We can plot f0 contours one by one and inspect the manually annotated / detected frequency jumps visually (not executed here):

```{r eval=FALSE}
check_pj(jumpThres = 20, jumpWindow = 200, plot = TRUE)
```

## Run optimization

```{r class.source = 'fold-hide'}
data_file = '../data/validate_pitchJump.RDS'
if (file.exists(data_file)) {
  out_pj = readRDS(data_file)
} else {
  out_pj = expand.grid(jumpThres = seq(.5, 24, .5),
                       jumpWindow = seq(20, 300, 10))
  n = nrow(out_pj)
  time_start = proc.time()
  for (i in 1:n) {
    temp = check_pj(jumpThres = out_pj$jumpThres[i], 
                    jumpWindow = out_pj$jumpWindow[i])
    # summary(temp$out[temp$out$n_manual > 0, ])
    out_pj[i, c('precision', 'recall', 'F1')] = temp[c('precision', 'recall', 'F1')]
    reportTime(i, time_start, n)
  }
  saveRDS(out_pj, data_file)
}
summary(out_pj)
```

Plot the results:

```{r}
ggplot(out_pj, aes(jumpThres, jumpWindow, fill = F1)) + geom_tile()
```

Look at the settings that provide the best F1 / recall / precision:

```{r}
head(out_pj[order(out_pj$F1, decreasing = TRUE), ])
head(out_pj[order(out_pj$precision, decreasing = TRUE), ])
head(out_pj[order(out_pj$recall, decreasing = TRUE), ])
```

Clearly, we can have nearly perfect recall if we treat even minor changes in f0 as frequency jumps, but precision is poor regardless of the settings - we can't really tell apart manually flagged pitch jumps vs. other rapid changes in f0 from one STFT frame to the next.

A nicer plot of optimization:

```{r class.source = 'fold-hide'}
d = reshape2::dcast(jumpThres ~ jumpWindow, data = out_pj, value.var = 'F1')
rownames(d) = d[, 1]
d = as.matrix(d[, -1])
mycolors = rev(rainbow(100) [1:90])
# png('../pix/pitchJump_validate.png', width = 15, height = 15, units = 'cm', res = 600)
soundgen:::filled.contour.mod(
  x = as.numeric(rownames(d)), 
  y = as.numeric(colnames(d)), 
  z = d, col = mycolors, nlevels = length(mycolors), 
  xlab = 'jumpThres', ylab = 'jumpWindow'
)
# dev.off()
```


```{r eval=FALSE, echo=FALSE}
# plot and save a legend
png('../pix/pitchJump_validate_legend.png', width = 5, height = 11, units = 'cm', res = 600)
nlevels = length(mycolors)
zlim = range(d)
levels = pretty(zlim, nlevels)
plot.new()
plot.window(xlim = c(0, 1), ylim = range(levels), xaxs = "i", yaxs = "i")
rect(0, levels[-length(levels)], 1, levels[-1L], col = mycolors, border = NA)
axis(4)
# box()
dev.off()
```

Final stats for a reasonable combination of parameters:

```{r}
# manual pitch tracking
df = readRDS('../data/bs_per_frame.RDS')
temp1 = check_pj(jumpThres = 20, jumpWindow = 100, tol_ms = 50)
temp1$precision  # % of the found jumps that are true positives
temp1$recall  # % of manually annotated pitch jumps that are successfully detected
summary(temp1$out)  # mean n_found = 0.1
```

```{r}
# automatic pitch tracking
df = readRDS('../data/bs_per_frame_autom.RDS')
temp2 = check_pj(jumpThres = 20, jumpWindow = 100, tol_ms = 50)
temp2$precision  # % of the found jumps that are true positives
temp2$recall  # % of manually annotated pitch jumps that are successfully detected
summary(temp2$out)  # n_found is way too high: mean = 1.4 (b/c of errors in automatic pitch tracking)
```


# Summary

We optimized the key parameters – the threshold rate of change (“jump threshold”) and the averaging window (“jump window”) – against manually annotated frequency jumps in our corpus of 1518 recordings of human nonverbal vocalizations, singing, and speech with manually verified fo contours, of which 65 recordings (8 speech samples and 57 nonverbal vocalizations) were judged to contain a total of 130 perceptually detectable frequency jumps 130 true positives is not much for formal optimization, and a simple grid search demonstrates that the best tested combination of control parameters (jump threshold = 19.5 octaves/s, jump window =  90 ms) achieves an F1-score of only 0.30, with a precision of 0.26 (proportion of manually annotated frequency jumps that are successfully detected) and a recall of 0.35 (proportion of detected frequency jumps that coincide with manual annotations).
