---
title: 'Checking inter-rater reliability of manual NLP annotation'
author: "Andrey Anikin"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 8
    fig_height: 6
    code_folding: show
---

```{r echo=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r echo=FALSE}
library(soundgen)
library(ggplot2)
```

The objective is to test how far expert raters agree about which NLP are present where in a sample of human speech, nonverbal vocalizations, and singing. 23 recordings were selected at random from a larger corpus (Anikin et al. 2023 "Beyond speech: exploring diversity in the human vocal space" iScience 26 (11): 108204). All 23 recordings were considered to contain NLP in the original publication; here we compare NLP annotations provided by 10 expert raters.

## Load all annotations

```{r}
ann = read.csv('../data/annot_anon.csv', stringsAsFactors = TRUE)
length(unique(ann$file))
table(ann$rater, ann$label)
```

Thus, we have 4 NLP types annotated in 23 recordings (2 to 10 s in duration) by `r length(unique(ann$rater))` human raters.


## Plot all annotations
```{r fig.width=12, fig.height=12, class.source = 'fold-hide'}
# png('../pix/annot_anon.png', width = 25, height = 30, units = 'cm', res = 300)
colors = list(
  'Frequency jump' = 'black',
  'Subharmonics' = 'blue',
  'Sidebands' = 'green',
  'Chaos' = 'red'
)
raters = as.character(unique(ann$rater))
nRaters = length(unique(ann$rater))

par(mfrow = c(6, 4), mar = c(2, 2, 4, 2))
for (file in unique(ann$file)) {
  temp = ann[ann$file == file, ]
  dur = temp$dur[1]
  # sort by NLP so arrows (FJ's) come last, over the segments
  temp = rbind(temp[temp$label != 'Frequency jump', ],
               temp[temp$label == 'Frequency jump', ])
  # par(mfrow = c(nRaters, 1))
  plot(0, 0, type = 'n', xlab = '', ylab = '', yaxt = 'n', bty = 'n', xlim = c(0, dur), ylim = c(0, nRaters + 1), main = file, cex.main = 0.5, cex.axis = .75)
  for (r in 1:nRaters) {
    rater = raters[r]
    temp2 = temp[temp$rater == rater, ]
    if (nrow(temp2) > 0) {
      for (a in 1:nrow(temp2)) {
        if (temp2$label[a] == 'Frequency jump') {
          arrows(x0 =  temp2$from[a], x1 =  temp2$from[a],
                 y0 = r + .6, y1 = r,
                 lwd = 2, length = .03)
        } else {
          segments(x0 = temp2$from[a], x1 = temp2$to[a],
                   y0 = r, y1 = r, col = colors[[temp2$label[a]]], lwd = 3)
        }
      }
    }
  }
}
# legend
plot(c(0, 5), c(0.5, 5.5), type = 'n', xlab = '', ylab = '', bty = 'n', xaxt = 'n', yaxt = 'n')
for (i in 1:length(colors)) {
  segments(x0 = 0, x1 = 2.5, y0 = i, y1 = i, col = colors[[i]], lwd = 10)
  text(5, i, label = names(colors)[i], adj = 1)
}
arrows(x0 = 0, x1 =  0, y0 = 5.5, y1 = 4.5, lwd = 2, length = .03)
text(5, 5, label = 'Frequency jump', adj = 1)
par(mfrow = c(1, 1))
# dev.off()
```


## Calculate concordance

```{r class.source = 'fold-hide'}
# status per 100 ms frame
step = .1
cc_type = cc_onoff = NULL
for (file in unique(ann$file)) {
  dur = ann$dur[ann$file == file] [1]
  temp_type = temp_onoff = as.data.frame(list(frame = 1:(dur / step), file = file))
  temp_type[, raters] = 'tonal'
  temp_onoff[, raters] = FALSE
  for (r in raters) {
    # any NLP with non-zero duration (ie all except frequency jumps)
    temp2 = ann[ann$file == file & ann$rater == r & ann$label != 'Frequency jump', ]
    if (nrow(temp2) > 0) {
      for (i in 1:nrow(temp2)) {
        idx = as.numeric(round(temp2[i, c('from', 'to')] / step))
        temp_type[idx[1]:idx[2], r] = temp2$label[i]
        temp_onoff[idx[1]:idx[2], r] = TRUE
      }
    }
  }
  cc_type = rbind(cc_type, temp_type)
  cc_onoff = rbind(cc_onoff, temp_onoff)
}

## average agreement between pairs of raters about NLP on/off
# all possible pairs of raters
pairs = expand.grid(raters, raters, stringsAsFactors = FALSE)
pairs = pairs[pairs[1] != pairs[2], ]
unique_pairs = idx_exclude = NULL
for (i in 1:nrow(pairs)) {
  pair_i = paste0(pairs[i, ], collapse = '_')
  pair_i_rev = paste0(rev(pairs[i, ]), collapse = '_')
  if (pair_i %in% unique_pairs | pair_i_rev %in% unique_pairs) {
    idx_exclude = c(idx_exclude, i)
  } else {
    unique_pairs = c(unique_pairs, pair_i)
  }
}
pairs = pairs[-idx_exclude, ]

for (i in 1:nrow(pairs)) {
  r1 = pairs$Var1[i]
  r2 = pairs$Var2[i]
  pairs$cor_onoff[i] = mean(c(
    mean(aggregate( (cc_onoff[, r1] == cc_onoff[, r2]) ~ cc_onoff[, r1], cc_onoff, mean)[, 2]),
    mean(aggregate( (cc_onoff[, r1] == cc_onoff[, r2]) ~ cc_onoff[, r2], cc_onoff, mean)[, 2])
  ))
  pairs$cor_type[i] = mean(c(
    mean(aggregate( (cc_type[, r1] == cc_type[, r2]) ~ cc_type[, r1], cc_type, mean)[, 2]),
    mean(aggregate( (cc_type[, r1] == cc_type[, r2]) ~ cc_type[, r2], cc_type, mean)[, 2])
  ))
}

## Frequency jumps
tol_ms = 100  # considered the same if within ±50 ms
for (i in 1:nrow(pairs)) {
  r1 = pairs$Var1[i]
  r2 = pairs$Var2[i]
  ann_i = ann[ann$rater %in% c(r1, r2) & ann$label == 'Frequency jump', ]
  ann_i$match = FALSE
  # ann_i = ann_i[order(ann_i$file), ]
  if (length(unique(ann_i$rater)) < 2) {
    # one rater didn't annotate any FJ's
    pairs$fj[i] = 0
  } else {
    for (r in 1:nrow(ann_i)) {
      idx = which(ann_i$file == ann_i$file[r] & 
                    ann_i$rater != ann_i$rater[r] &
                    abs(ann_i$from - ann_i$from[r]) * 1000 < tol_ms)
      if (length(idx) > 0) ann_i$match[r] = TRUE
    }
    pairs$fj[i] = mean(ann_i$match)
  }
}
# pairs
colMeans(pairs[, 3:5])
```

cor_noff: average agreement between pairs of raters regarding whether or not a 100-ms frame contains any continuous NLP (sidebands, subharmonics, or chaos)

cor_type: average agreement between pairs of raters regarding the type of NLP (none / sidebands / subharmonics / chaos) in each 100-ms frame

fj: average agreement between pairs of raters regarding the presence of a frequency jump within ±50 ms of an annotated one


