---
title: 'NLP synthesis: AM/FM/subharmonics'
author: "Andrey Anikin"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 8
    fig_height: 6
---

```{r echo=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r echo=FALSE}
library(soundgen)
```

# Amplitude modulation (AM)

## AM in synthetic sounds

- Multiply sound by a slowly oscillating signal

- Sinusoidal AM creates a single pair of extra harmonics around each f0 harmonic

- Non-sinusoidal AM creates sidebands

- Pitch-synchronous AM creates subharmonics

- see ?soundgen::addAM for details


```{r}
s1_am = soundgen(
  sylLen = 700,
  pitch = list(time = c(0, .15, .3, .9, 1),
               value = c(140, 360, 365, 245, 120)),
  rolloff = c(-5, -8),
  formants = 'uoo0',
  noise = c(-15, -20),
  amFreq = 75, amDep = c(0, 0, 75, 55, 0, 0),
  plot = TRUE, yScale = 'bark', heights = c(1, 1))
# playme(s1_am)
```


```{r echo=FALSE}
seewave::savewav(s1_am, 16000, filename = 'ex_audio/ex1_am.wav')
```

<audio controls> <source src="ex_audio/ex1_am.wav"> </audio>


## Sinusoidal AM

```{r}
soundgen(pitch = 900, amFreq = 100, amDep = 50, amType = 'sine', 
         plot = TRUE, ylim = c(0, 4), heights = c(1, 1))
```


## Non-sinusoidal AM

```{r}
soundgen(pitch = 900, amFreq = 100, amDep = 50, amShape = .25,
         plot = TRUE, ylim = c(0, 4), heights = c(1, 1))
```


## Add AM to recording

```{r echo=FALSE}
spectrogram('ex_audio/speechEx.wav', yScale = 'bark')
```

<audio controls> <source src="ex_audio/speechEx.wav"> </audio>


```{r}
sp_am20Hz = addAM('ex_audio/speechEx.wav', 
                  amFreq = 20, 
                  amDep = c(0, 100))
spectrogram(sp_am20Hz, 16000, yScale = 'bark', heights = c(2, 1))
```

```{r echo=FALSE}
seewave::savewav(sp_am20Hz, 16000, filename = 'ex_audio/speechEx_am20.wav')
```

<audio controls> <source src="ex_audio/speechEx_am20.wav"> </audio>



# Frequency modulation (FM) 

## FM in synthetic sounds

Regular, deterministic vibrato (the depth and frequency changes over time, but executing the command always produces nearly the same result):

```{r}
fm_rv = soundgen(
  vibratoDep = 0:3, vibratoFreq = 7:5, 
  sylLen = 2000, pitch = c(300, 280), 
  plot = TRUE, ylim = c(0, 3))
```


```{r echo=FALSE}
seewave::savewav(fm_rv, 16000, filename = 'ex_audio/fm_rv.wav')
```

<audio controls> <source src="ex_audio/fm_rv.wav"> </audio>


Rapid sinusidal FM produces multiple sidebands, which looks and sounds very similar to (nonsinusoidal) AM:

```{r}
fm_rapid = soundgen(
  vibratoDep = 2, vibratoFreq = 150,
  sylLen = 2000, pitch = 1000, 
  plot = TRUE, ylim = c(0, 3), invalidArgAction = 'ignore')
```

```{r echo=FALSE}
seewave::savewav(fm_rapid, 16000, filename = 'ex_audio/fm_rapid.wav')
```

<audio controls> <source src="ex_audio/fm_rapid.wav"> </audio>


Compare to the same sound with AM instead of FM at 150 Hz:

```{r}
am_rapid = soundgen(
  amDep = 40, amFreq = 150, amShape = -0.5,
  sylLen = 2000, pitch = 1000, 
  plot = TRUE, ylim = c(0, 3))
```

```{r echo=FALSE}
seewave::savewav(am_rapid, 16000, filename = 'ex_audio/am_rapid.wav')
```

<audio controls> <source src="ex_audio/am_rapid.wav"> </audio>



## Add FM to a recording

```{r}
fmFreq = 10  # Hz
fmDep = 1.2  # semitones
fmDep_linear = (2 ^ (fmDep / 12) - 1) / 2
target = tuneR::readWave('ex_audio/speechEx.wav')
multPitch = cos(2 * pi * fmFreq * (1:length(target@left)) / target@samp.rate) * fmDep_linear + 1
plot(multPitch, type = 'l')
# verify that multiplying f0 by multPitch produces FM of ±fmDep/2 semitones:
log2(range(multPitch)) * 12
```

```{r}
sp_fm10Hz = shiftPitch(target, multPitch = multPitch)
spectrogram(sp_fm10Hz, 16000, yScale = 'bark', heights = c(2, 1))
# playme(sp_fm10Hz, 16000)
```

```{r echo=FALSE}
seewave::savewav(sp_fm10Hz, 16000, filename = 'ex_audio/speechEx_fm10.wav')
```

<audio controls> <source src="ex_audio/speechEx_fm10.wav"> </audio>



# Subharmonics

## Fully synthetic

- Literally add new sine waves (partials) between f0 harmonics

- subRatio controls f0/g0

- subDep controls the amplitude of added partials relative to f0


```{r}
s_sh = soundgen(
  sylLen = 1200,
  pitch = list(time = c(0, .15, .3, .9, 1),
               value = c(140, 360, 365, 245, 120)),
  rolloff = c(-5, -8),
  formants = 'uoo0',
  noise = c(-15, -20),
  subRatio = 2, subDep = c(0, 0, 0, 35, 15))
spectrogram(s_sh, 16000, yScale = 'bark', heights = c(3, 2))
```

```{r echo=FALSE}
seewave::savewav(s_sh, 16000, filename = 'ex_audio/s_sh.wav')
```

<audio controls> <source src="ex_audio/s_sh.wav"> </audio>


## Resynthesis / manipulation

Subharmonics can be imitated by adding pitch-synchronous AM to an existing recording:


```{r echo=FALSE}
pitch = read.csv('ex_audio/speechEx_pitch.csv')$pitch  # may require checking manually for best results
pitch = eval(parse(text = paste0('c(', pitch, ')')))
plot(1:length(pitch) * 25, pitch, type = 'l', xlab = 'Time, ms', ylab = 'Pitch, Hz')
```

```{r}
sp_amPS = addAM('ex_audio/speechEx.wav', 
                amFreq = pitch / 2, 
                amDep = c(25, 75, 25))
spectrogram(sp_amPS, 16000, ylim = c(0, 1))
```

```{r echo=FALSE}
seewave::savewav(sp_amPS, 16000, filename = 'ex_audio/speechEx_amPS.wav')
```

<audio controls> <source src="ex_audio/speechEx_amPS.wav"> </audio>

