---
title: 'NLP synthesis: biphonation'
author: "Andrey Anikin"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 8
    fig_height: 6
---


```{r echo=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r echo=FALSE}
library(soundgen)
```



## Biphonation

- Two sound sources in parallel that don't infuence each other's airflow --> additive effects, two independent sources

- Two sound sources sitting one on top of the other in the airway or strongly coupled --> multiplicative effects, two interacting sources


## Parallel

Like two notes played simultaneously on a piano: we can just synthesize them independently and add up the results:

```{r echo=FALSE}
par(mfrow = c(1, 3))
s1 = cos(2 * pi * 20 * (1:1000)/1000)
plot(s1, type = 'l')
s2 = cos(2 * pi * 2 * (1:1000)/1000)
plot(s2, type = 'l')
plot(s1 + s2, type = 'l')
```

The spectrum shows two frequencies (here two pure tones):

```{r echo=FALSE}
seewave::spec(rep(s1 + s2, 10), f = 1000, dB = 'max0', flim = c(0, 0.1))
```


## In series

One source "downstream" of another, modulating baseline air pressure

```{r echo=FALSE}
par(mfrow = c(1, 3))
# the "higher" frequency ("higher" in Hz and anatomically, in the sense that
# it's the one sitting downstream and being modulated)
s1 = cos(2 * pi * 20 * (1:1000)/1000)
plot(s1, type = 'l')
# the "lower" frequency modulating the first one (rescaled to range from 0 to 1)
s2 = (cos(2 * pi * 2 * (1:1000)/1000) + 1) / 2
range(s2)
plot(s2, type = 'l')
plot(s1 * s2, type = 'l')
```

The spectrum shows the higher frequency modulated by the lower one:

```{r echo=FALSE}
seewave::spec(rep(s1 * s2, 10), f = 1000, dB = 'max0', flim = c(0, 0.1))
```


Implications for biphonation:

- Create two sources separately

- Add with a multiplicative term if one modulates the other



## Example of biphonation: whistle & tone

```{r}
s1 = soundgen(pitch = c(200), sylLen = 500, rolloff = -15, 
              addSilence = 0, attackLen = 5, samplingRate = 40000)
s1 = soundgen:::zeroOne(s1)
spectrogram(s1, 40000, ylim = c(0, 5), heights = c(2, 1), osc = FALSE)
```


```{r echo=FALSE}
seewave::savewav(s1, 40000, filename = 'ex_audio/biphon_tone.wav')
```

<audio controls> <source src="ex_audio/biphon_tone.wav"> </audio>


```{r}
s2 = soundgen(pitch = c(2500, 2200, 2700), sylLen = 500, rolloff = -30, 
              jitterDep = .25, shimmerDep = 25, addSilence = 0, 
              attackLen = 5, formants = NULL, samplingRate = 40000)
s2 = s2[1:length(s1)]
spectrogram(s2, 40000, ylim = c(0, 10), windowLength = 15, osc = FALSE)
```


```{r echo=FALSE}
seewave::savewav(s2, 40000, filename = 'ex_audio/biphon_whistle.wav')
```

<audio controls> <source src="ex_audio/biphon_whistle.wav"> </audio>


Whistle & tone added:

```{r}
spectrogram(s1 + s2, 40000, ylim = c(0, 10), windowLength = 15, osc = FALSE)
```


```{r echo=FALSE}
seewave::savewav(s1 + s2, 40000, filename = 'ex_audio/biphon_add.wav')
```

<audio controls> <source src="ex_audio/biphon_add.wav"> </audio>


Whistle & tone multiplied:

```{r echo=FALSE, fig.width=8, fig.height=5}
spectrogram(s1 * s2, 40000, ylim = c(0, 10), windowLength = 15, osc = FALSE)
```



```{r echo=FALSE}
seewave::savewav(s1 * s2, 40000, filename = 'ex_audio/biphon_mult.wav')
```

<audio controls> <source src="ex_audio/biphon_mult.wav"> </audio>


## Example of biphonation: synthetic wapiti

The low-freq component:
```{r}
formants = c(130, 400, 600, 800, 1050, 1250, 1370, 1580)
s1 = soundgen(
  sylLen = 1450,
  attackLen = c(150, 50),
  pitch = c(120, 150, 145, 120),
  jitterDep = .25, shimmerDep = 10,
  rolloff = -20,
  formants = formants,
  formantDep = 2, formantWidth = 0.5, formantCeiling = 5,
  noise = c(-20, -10),
  temperature = .02,
  addSilence = 0,
  samplingRate = 22050,
  pitchSamplingRate = 22050,
  plot = F, ylim = c(0, 5)
)
```

The high-frequency component:

```{r}
s2 = soundgen(
  sylLen = 2900,
  attackLen = 200,
  pitch = c(670, 1000, 1102, 1239, 1367, 1376, 1599, 1579, 1559, 1314),
  jitterDep = c(.05, .05, .2, .2, .05), shimmerDep = 5,
  rolloff = c(-10, -2, 0, -4),
  noise = c(-40, -10),
  formants = formants,
  formantDep = 1, formantWidth = 0.5,
  formantLocking = c(.5, rep(1, 8)),
  formantCeiling = 5,
  temperature = 1e-6,
  addSilence = 0,
  samplingRate = 22050,
  pitchSamplingRate = 22050,
  play = F, plot = T, osc = T, yScale = 'bark'
)
```

Add them up with modulation:

```{r}
# make sure both are the same length (g0 ends 150 ms before f0)
s1n = soundgen:::matchLengths(s1, len = length(s2) + .13 * 22050, padDir = 'left')  
s1_01 = soundgen:::zeroOne(s1n) # set range to [0, 1]
s2n = soundgen:::matchLengths(s2, len = length(s1n), padDir = 'right') 
s = s1n/4 + s2n * s1_01
spectrogram(s, 22050, yScale = 'bark')
# playme(s, 22050)
```


```{r echo=FALSE}
seewave::savewav(s, 22050, filename = 'ex_audio/biphon_bugle.wav')
```

<audio controls> <source src="ex_audio/biphon_bugle.wav"> </audio>

